Understanding School Shootings: Analyzing Trends, Demographics, and Geographic Factors Using Bayesian Methods

### 2024-12-15 paper/paper.qmd
update the typo and format
update the typo and format

---
title: "Understanding School Shootings: Analyzing Trends, Demographics, and Geographic Factors Using Bayesian Methods"
subtitle: "A Data-Driven Investigation into the Predictors of School Shootings in the United States (1999–2024)"
author: 
  - Ruizi Liu
thanks: "Code and data are available at: [https://github.com/RIRI0527/school_shooting_analysis.git](https://github.com/RIRI0527/school_shooting_analysis.git)."
date: today
date-format: long
abstract: "School shootings remain a critical issue in the United States, with far-reaching social and psychological consequences. This study investigates the factors influencing the occurrence of school shootings, focusing on temporal trends, school type, demographic composition, and geographic context. Using a Bayesian logistic regression model, this analysis incorporates both observed and simulated data to explore the relationship between these predictors and the likelihood of a shooting. The results reveal significant associations between the predictors and school shootings. Temporal trends indicate an alarming increase in incidents over time, while public schools and urban locales exhibit higher risks compared to private schools and rural areas. Demographic composition also plays a role, with schools characterized by specific racial majorities showing distinct patterns in the likelihood of shootings. This report emphasizes the need for targeted, data-driven interventions that consider institutional and geographic variations in risk. By leveraging probabilistic modeling, this study provides actionable insights for policymakers aiming to mitigate the occurrence of school shootings and improve the safety of educational environments across the United States. "
format: pdf
number-sections: true
bibliography: references.bib
editor: 
  markdown: 
    wrap: sentence
---

```{r}
#| include: false
#### Installing the packages ####
required_packages <- c("tidyverse", "here", "maps")
for (p in required_packages) {
  if (!require(p, character.only = TRUE)) {
    install.packages(p, dependencies = TRUE)
    library(p, character.only = TRUE)
  }
}
```

@@

The measurement process involved triangulating information from:
- **News Reports**: Verified through multiple outlets for accuracy.
- **Law Enforcement Records**: Provided timelines, locations, and key details.
- **School Data**: Publicly available records enriched the dataset with demographic and institutional contexts.

@@

The dataset’s strengths lie in its:
- **Comprehensiveness**: Captures a wide range of school shootings with detailed contexts.
- **Transparency**: Relies on well-documented and publicly accessible sources.
- **Relevance**: Includes key variables such as urbanicity and demographics critical to understanding school shootings.

@@

We used the`R`programming language [@citeR], the `here` package [],
the `dplyr` package [@R_dplyr], the `tidyr` package [@R_tidyr], the `maps` package [@R_maps], the `ggplot2` package [@R_ggplot2], and the`scale` package [@R_scales] to clean the data. We used the `tidyverse` package [@tidyverse], the `maps` package [@maps], the `ggplot2` package [@R_ggplot2], the `knitr` package [@knitr], the `kableExtra` package [@kableExtra] to plot the dataset and the visualization.

@@

In addition, because the original data only recorded the shootings that occurred and not the ones that did not, but we wanted to simulate the logistic regression model, we generated synthetic data based on the cleaned dataset to reflect the data that did not occur. The main components that were simulated were latitude, longitude, school ethnicity, lunch, and staffing, and to ensure that the data were reasonable, the synthetic data were generated within the allowable range of variability (a reasonable float between the maximum and minimum of the variable), and to ensure that the model was implementable, the synthetic data were generated within the allowable range of variation.Synthetic data and raw data both have 415 observations.at the same time, we made a distinction between synthetic data and raw data inside the dataset by labelling their data source in data_source and creating a new binary variable to indicate whether a shooting occurred or not.

@@

```{r}
#| label: tlb-data-overview
#| tib-cap: The specific columns in the dataset
#| echo: false
#| warning: false

@@

# Create the kable table
kable(summary_data, 
      col.names = c("Year", "State", "School Type", "Top 1 Race", "Ulocale", "Data Source", "School Shooting"),
      caption = "Summary of School Shooting Data") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                full_width = F, 
                position = "left")
```

@@

#### Predictors
- **Year**: Captures temporal trends in shootings.
- **School Type**: Public vs. private schools.
- **Urban Locale**: Urbanicity levels from NCES data.
- **Enrollment**: Total number of students in the school.
- **Dominant Race**: Primary racial group in the school.

#### Priors
- Intercept $\beta_0\$: $\text{Normal}(0, 5)$, reflecting baseline uncertainty.
- Coefficients $\beta_k$: $\text{Normal}(0, 2.5)$, allowing moderate variability.

@@

```{r tbl-model-summary, echo=FALSE, warning=FALSE, startup_message = FALSE}

#| fig-cap: Bayesian Logistic Regression Coefficients

# config_modelsummary(startup_message = FALSE)
library(modelsummary)
model <- readRDS(file = here::here("models/model_bayes.rds"))

modelsummary(model, 
             output = "kableExtra",
             stars = FALSE,
             statistic = "std.error",
             title = "Bayesian Logistic Regression Coefficients")
```

@@

## Diagnostics

Model diagnostics confirmed the robustness of the Bayesian logistic regression.

@@

1.  **Posterior Predictive Check** @fig-ppc: The posterior predictive distribution aligns closely with the observed data, indicating a good model fit.

2.  **Convergence**: @fig-trace-plots displays trace plots for key parameters, showing well-mixed chains that reached stationarity. All parameters had ($\hat{R} < 1.1$), as shown in @fig-rhat.

3.  **Credibility Intervals** @fig-credible-intervals: The $90%$ credibility intervals for each parameter are plotted, highlighting significant predictors (intervals that exclude 0).

>>>>

---
title: "Understanding School Shootings: Analyzing Trends, Demographics, and Geographic Factors Using Bayesian Methods"
subtitle: "A Data-Driven Investigation into the Predictors of School Shootings in the United States (1999–2024)"
author: 
  - Ruizi Liu
thanks: "Code and data are available at: [https://github.com/RIRI0527/school_shooting_analysis.git](https://github.com/RIRI0527/school_shooting_analysis.git)."
date: today
date-format: long
abstract: "School shootings remain a critical issue in the United States, with far-reaching social and psychological consequences. This study investigates the factors influencing the occurrence of school shootings, focusing on temporal trends, school type, demographic composition, and geographic context. Using a Bayesian logistic regression model, this analysis incorporates both observed and simulated data to explore the relationship between these predictors and the likelihood of a shooting. The results reveal significant associations between the predictors and school shootings. Temporal trends indicate an alarming increase in incidents over time, while public schools and urban locales exhibit higher risks compared to private schools and rural areas. Demographic composition also plays a role, with schools characterized by specific racial majorities showing distinct patterns in the likelihood of shootings. This report emphasizes the need for targeted, data-driven interventions that consider institutional and geographic variations in risk. By leveraging probabilistic modeling, this study provides actionable insights for policymakers aiming to mitigate the occurrence of school shootings and improve the safety of educational environments across the United States. "
format: pdf
prefer-html: true
number-sections: true
bibliography: references.bib
editor: 
  markdown: 
    wrap: sentence
---

@@

The measurement process involved triangulating information from:

- **News Reports**: Verified through multiple outlets for accuracy.

- **Law Enforcement Records**: Provided timelines, locations, and key details.

- **School Data**: Publicly available records enriched the dataset with demographic and institutional contexts.

@@

The dataset’s strengths lie in its:

- **Comprehensiveness**: Captures a wide range of school shootings with detailed contexts.

- **Transparency**: Relies on well-documented and publicly accessible sources.

- **Relevance**: Includes key variables such as urbanicity and demographics critical to understanding school shootings.

@@

We used the`R`programming language [@citeR], the `here` package [@here],the `dplyr` package [@R_dplyr], the `tidyr` package [@R_tidyr], the `maps` package [@R_maps], the `ggplot2` package [@R_ggplot2], and the`scale` package [@R_scales] to clean the data. We used the `tidyverse` package [@tidyverse], the `maps` package [@maps], the `ggplot2` package [@R_ggplot2], the `knitr` package [@knitr], the `kableExtra` package [@kableExtra] to plot the dataset and the visualization.

@@

In addition, because the original data only recorded the shootings that occurred and not the ones that did not, but we wanted to simulate the logistic regression model, we generated synthetic data based on the cleaned dataset to reflect the data that did not occur. The main components that were simulated were latitude, longitude, school ethnicity, lunch, and staffing, and to ensure that the data were reasonable, the synthetic data were generated within the allowable range of variability (a reasonable float between the maximum and minimum of the variable), and to ensure that the model was implementable, the synthetic data were generated within the allowable range of variation.Synthetic data and raw data both have 415 observations.at the same time, we made a distinction between synthetic data and raw data inside the datasets by labeling their data source in data_source and creating a new binary variable to indicate whether a shooting occurred or not.

@@

```{r}
#| label: tbl-data-overview
#| tbl-cap: The selected columns in the dataset
#| echo: false
#| warning: false

@@

#### Generate the data table ####
summary_data %>%
  head(10) %>%
  kable(
    format = "html",
    caption = "Summary of School Shooting Data",
    col.names = c("Year", "State", "School Type", "Top 1 Race", "Ulocale", "Data Source", "School Shooting")  # Ensure this matches column count
  ) %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>%
  column_spec(1, bold = TRUE, color = "blue")
```

@@

#### Predictors

- **Year**: Captures temporal trends in shootings.

- **School Type**: Public vs. private schools.

- **Urban Locale**: Urbanicity levels from NCES data.

- **Enrollment**: Total number of students in the school.

- **Dominant Race**: Primary racial group in the school.

#### Priors
- Intercept $\beta_0$: $\text{Normal}(0, 5)$, reflecting baseline uncertainty.
- Coefficients $\beta_k$: $\text{Normal}(0, 2.5)$, allowing moderate variability.

@@

```{r tbl-model-summary, echo=FALSE, warning=FALSE, startup_message = FALSE}
#| tbl-cap: "Bayesian Logistic Regression Coefficients"
#| label: tbl-model-summary

library(modelsummary)

# Render the model summary as a LaTeX-compatible table
modelsummary(
  model, 
  output = "latex",
  stars = FALSE,
  statistic = "std.error",
  title = "Bayesian Logistic Regression Coefficients"
)

@@

## Diagnostics
Model diagnostics confirmed the robustness of the Bayesian logistic regression.

@@

1.  **Posterior Predictive Check** [@fig-ppc]: The posterior predictive distribution aligns closely with the observed data, indicating a good model fit.

2.  **Convergence** [@fig-trace-plots]: displays trace plots for key parameters, showing well-mixed chains that reached stationarity. All parameters had ($\hat{R} < 1.1$), as shown in @fig-rhat.

3.  **Credibility Intervals** [@fig-credible-intervals]: The $90%$ credibility intervals for each parameter are plotted, highlighting significant predictors (intervals that exclude 0).
